---
title: "AD_Phenotyping"
author: "KBazemore"
date: "2024-09-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(root.dir = "/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/pathPRS-UKBB/phenotype")
```

Loading libraries
```{r}
library(dplyr)
library(gtsummary)
library(cardx)
library(gt)
```


UKBB Participant Inclusion/Exclusion

Input files
```{r}
#Phenotype file eg sex, PCs
phenotypes <- read.csv("/home/bazemork/Covariates/phenotypes.csv", header = T)

#Enviro covariates file 
enviro <- read.csv("/home/bazemork/Covariates/enviro.csv", header = T)

#AD Pheno and popdense variable
exp_out <- read.table("/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/EPID6210_Final/popdense_covar.txt", header = T)


#Personal history of AD (true cases) file
#ad_eids <- read.csv("/home/bazemork/Covariates/EIDs_ALL_AD.csv", header = T)

withdrawals <- read.csv("/home/bazemork/UKB_Withdrawals/ids_to_remove_nodups01112024.csv", header = T)
```


Applying inclusion/exclusion criteria

Step 1
```{r}
int1 <- dplyr::filter(phenotypes, genetic_grouping ==1)


print(paste0("Number of UKBB participants with genetic similarity to European reference populations: ", nrow(int1), "  Number of participants with genetic similarity to non-European reference populations: ", nrow(phenotypes)-nrow(int1)))
rm(phenotypes)
```
Step 2
```{r}
#Steps 2a - 2d
int2a <- dplyr::filter(int1, is.na(outlier_het_miss))
print(paste0("Outliers for heterozygosity and missingness: ", nrow(int1)-nrow(int2a)))
rm(int1)

int2b <- dplyr::filter(int2a, kinship !=10)
print(paste0("Relatedness: ", nrow(int2a)-nrow(int2b)))
rm(int2a)

int2c <- dplyr::filter(int2b, is.na(sex_aneuploidy))
print(paste0("Sex chromosome aneuploidy: ", nrow(int2b)-nrow(int2c)))
rm(int2b)

int2d <- dplyr::filter(int2c, inpca ==1)
print(paste0("Not included in PCA: ", nrow(int2c)-nrow(int2d)))
rm(int2c)
```

Step 3
```{r}
#Remove withdrawals
int3 <- dplyr::filter(int2d, !eid %in% withdrawals$x) 

print(paste0("Withdrawals removed: ", nrow(int2d)-nrow(int3)))

print(paste0("Total passed QC and in PCA and did not withdraw: ", nrow(int3)))

rm(withdrawals, int2d)
```

Step 4
```{r}
#Inclusion of participants dwelling in England/Wales in top four pop density categories
top_cats <- c(5, 6, 7, 8)
EW_topCat <- exp_out %>% dplyr::filter(f.20118.0.0 %in% top_cats)

int4 <- int3 %>% dplyr::filter(eid %in% EW_topCat$f.eid)
print(paste0("Exclusion due to non-top pop density categories: ", nrow(int3)-nrow(int4)))
rm(top_cats, int3, exp_out)
```

Step 5
```{r}
#Inclusion of participants 60 and over
int5 <- int4 %>% dplyr::filter(participant_age_recriutment >= 60)
print(paste0("Exclusion of those under 60: ", nrow(int4)-nrow(int5)))
rm(int4)
```

AD Phenotype
```{r}
ad_codes <- c(11, 12, 21, 22)
EW_topCat2 <- EW_topCat %>% 
  dplyr::mutate(AD = case_when(f.42021.0.0 %in% ad_codes ~ 1, TRUE ~ 0)) %>% 
  dplyr::rename(eid = f.eid, pop_cat = f.20118.0.0) %>%
  dplyr::select(eid, pop_cat, AD)

int6 <- dplyr::left_join(int5, EW_topCat2) %>%
  dplyr::select(eid, AD, pop_cat, participant_age_recriutment, sex, current_smoking, past_smoking, etoh_intake, etoh_10yrsago, pc1, pc2, pc3, pc4)
rm(int5, EW_topCat, EW_topCat2)
```

Other Environmental Variables
```{r}
int7 <- dplyr::left_join(int6, enviro)
rm(int6, enviro)
```

Adding in information on APOE haplotype
```{r}
#Reading in data on APOE marker variants
APOE_training <- read.table("/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/Kunkle_pathPRS/recoded_training_APOE2.raw", header = T) %>% 
  dplyr::select(IID, rs429358_T, rs7412_C)

APOE_testing <- read.table("/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/Kunkle_pathPRS/recoded_testing_APOE2.raw", header = T) %>% 
  dplyr::select(IID, rs429358_T, rs7412_C)

#Merging data and recoding for APOE (haplotype) alleles
APOE <- rbind(APOE_training, APOE_testing) %>% 
  dplyr::mutate(
    rs429358 = case_when(rs429358_T == 0 ~ "CC", 
                         rs429358_T == 1 ~ "CT", 
                         rs429358_T == 2 ~ "TT"), 
    rs7412 = case_when(rs7412_C == 0 ~ "TT", 
                       rs7412_C == 1 ~ "CT", 
                       rs7412_C == 2 ~ "CC"),
    eid = IID) %>%
  dplyr::mutate(
    apoe1 = case_when(rs429358 == "CC" & rs7412 == "CC" ~ 4, 
                      rs429358 == "CT" & rs7412 == "CC" ~ 4, 
                      rs429358 == "TT" & rs7412 == "CC" ~ 3, 
                      rs429358 == "CT" & rs7412 == "CT" ~ 2, 
                      rs429358 == "TT" & rs7412 == "CT" ~ 2, 
                      rs429358 == "TT" & rs7412 == "TT" ~ 2), 
    apoe2 = case_when(rs429358 == "CC" & rs7412 == "CC" ~ 4, 
                      rs429358 == "CT" & rs7412 == "CC" ~ 3, 
                      rs429358 == "TT" & rs7412 == "CC" ~ 3, 
                      rs429358 == "CT" & rs7412 == "CT" ~ 4, 
                      rs429358 == "TT" & rs7412 == "CT" ~ 3, 
                      rs429358 == "TT" & rs7412 == "TT" ~ 2)) %>%
  dplyr::mutate(e4_count = case_when(apoe1 == 4 & apoe2 == 4 ~ 2, 
                                     apoe1 !=4 & apoe2 == 4 ~ 1, 
                                     apoe1 == 4 & apoe2 != 4 ~ 1, 
                                     apoe1 != 4 & apoe2 != 4 ~ 0)) %>%
  dplyr::select(eid, e4_count, apoe1, apoe2)

#Table of e4 allele count 
table(APOE$e4_count)

#Table of APOE allele 1 vs allele 2
table(APOE$apoe1, APOE$apoe2)

#Merging APOE data with cohort phenotype data
int8 <- merge(APOE, int7, by = "eid")
print(paste0("Exclusion of those without APOE data: ", nrow(int7)-nrow(int8)))
rm(int7, APOE, APOE_testing, APOE_training)
```

Saving UKBB phenotype data post inclusion/exclusion, phenotyping and APOE haplotypes (still need to add PRS exposure)
```{r}
save(int8, file = "/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/EPID6210_Final/included_cohort.rds", quote = F, row.names = F)
```




Demographics

Recoding outcome for demographics table
```{r}
whole.set <- rbind(training, testing)

whole.set <- whole.set %>% dplyr::mutate(outcome = case_when(control == "Yes" ~ "Control",
                                                                    case == "Yes" ~ "Case",
                                                                    proxy == "Yes" ~ "Proxy Case"))
```

Making demographics table
```{r}
new_label_overall <- paste0("Overall", "  \n", "N = 332,068")
new_label_training <- paste0("Training", "  \n", "N = 248,984")
new_label_testing <- paste0("Testing", "  \n", "N = 83,084")

whole.set <- whole.set %>% dplyr::mutate(in_set = case_when(eid %in% training$eid ~ "training", 
                                                            eid %in% testing$eid ~ "testing"))

saveRDS(whole.set, "/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/pathPRS-UKBB/phenotype/data/final_cohort_phenotypes")
whole.set <- readRDS("/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/pathPRS-UKBB/phenotype/data/final_cohort_phenotypes")

whole.set <- whole.set %>%
  dplyr::mutate(outcome_by2 = case_when(outcome == "Case" ~ "Case/Proxy",
                                        outcome == "Proxy Case" ~ "Case/Proxy",
                                        outcome == "Control" ~ "Control"))

demographics <- whole.set %>% 
  select(outcome_by2, participant_age_recriutment, sex, e4_count, maternal_history, paternal_history, in_set) %>% 
  tbl_summary(by = in_set, label = list(
    outcome_by2 ~ "AD Status",
    maternal_history ~ "Maternal dementia", 
    paternal_history ~ "Paternal dementia", 
    participant_age_recriutment ~ "Age at recruitment", 
    sex ~ "Male sex",
    e4_count ~ "APOE \u03B54 allele count"),
    missing = "no") %>% 
  add_overall() %>% 
  add_difference(everything() ~ "smd") %>% 
  modify_column_hide(columns = c(conf.low, conf.high)) %>%
  modify_header(stat_0 ~"{new_label_overall}", stat_1 ~"{new_label_testing}", stat_2 ~"{new_label_training}", estimate ~"SMD") %>%
  modify_table_body(~.x %>% dplyr::relocate(stat_1, .after = stat_2)) %>%
  as_gt() %>%
  cols_width(c(label) ~ px(250), everything() ~ px(115)) %>%
  tab_options(
    column_labels.font.weight = "bold", data_row.padding = px(1)
  ) %>%
  text_transform(
    locations = cells_body(columns = label, rows = 6),
    fn = function(x) paste0("<i>APOE</i> ε4 allele count")) 

demographics

saveRDS(demographics, file = "/s3buckets/adamnaj-lab-psom-s3-bucket-01/bazemork/pathPRS-UKBB/phenotype/data/demographics_traintest")
  #unable to run gtsave in server, download RDS to local machine and run gtsave there

```
